---
title: "Readme"
output: html_document
date: "2025-01-06"
---

# jointCompRisk

## Overview
The jointCompRisk R package provides a suite of competing risks analysis tools for use in clinical research or other fields where multiple event types are of interest. In particular, this package includes:

**Part I**: Standard Cumulative Incidence Function (CIF) inference for competing events.
**Part II**: Weighted CIF methods for incorporating weighted survival times, allowing different event types (e.g., discharge vs. death) to have distinct clinical interpretations.

To demonstrate the workflow without disclosing private data, jointCompRisk also ships with two dummy datasets (`dummy_actt_ch` and `dummy_actt_long`) and a data preparation function `prepare_actt_data()` that mimics the real cleaning and merging steps used in clinical trial analyses.

## Key Features
**1. Standard CIF Analysis**

`auc.func()`: Compute area under a CIF curve.
`auc.var.joint()`: Calculate an estimate of a combined measure Ψ and its variance.
`table1()`: Generate a 3x4 matrix comparing two groups (treatment vs. control) for CIF-based measures.

**2. Weighted CIF Analysis**
`w_surv()`: Core function for computing the weighted survival function.
`auc.ws()` & `sd.auc.ws()`: Calculate the weighted area under the curve and its standard error.
`table_wr()`: Summarize weighted AUC metrics (group difference, confidence intervals, p-values).

**3. Data Preparation and Dummy Data**
`prepare_actt_data()`: Illustrates how to clean & merge data into the competing risks format for both CIF and weighted CIF.
`dummy_actt_ch`, `dummy_actt_long`: Synthetic data frames for demonstration.

## Installation
You can install `jointCompRisk` directly from GitHub using devtools:
```{r}
install.packages("devtools")
devtools::install_github("cathyzzzhang/jointCompRisk")

library(jointCompRisk)
```

## Getting Started
#### Load and Explore Dummy Data
```{r}
# The package includes dummy data for illustration
data("dummy_actt_ch")
data("dummy_actt_long")

head(dummy_actt_ch)
head(dummy_actt_long)
```
These data frames have synthetic patient-level records resembling real-world trial data.

#### Prepare Data for Analysis
`prepare_actt_data()` demonstrates how the package merges and cleans data for standard and weighted CIF. In a real scenario, you would adapt your own data to have similar columns (TTRECOV, TTDEATH, etime, wU, etc.) before calling the analysis functions.
```{r}
# Creates a list of data frames for Part I and Part II analyses
res_list <- prepare_actt_data()
names(res_list) 
```
`res_list$data.w` is for standard CIF analysis (Part I), 
while `res_list$data.ws.death` and `res_list$data.ws.discharge` are for weighted CIF (Part II). Also, subsets like trt1, trt0 (treatment vs. control) are included.

#### Part I: Standard CIF Analysis
##### Table for CIF-Based Measures(`table1`)

You can estimate and compare cumulative incidence-based measures between two groups (e.g., trt1 and trt0) using table1(). For instance:

```{r}
res_table <- table1(
  data1 = res_list$trt1,
  data2 = res_list$trt0,
  tau   = 15,
  a     = 0,  
  b     = 1, 
  c     = 0 
)
res_table
```

This outputs a 3×4 matrix:

Row 1: Group 1 estimate (with 95% CI).
Row 2: Group 2 estimate (with 95% CI).
Row 3: Absolute difference and p-value.

#### Part II: Weighted CIF Analysis 
##### Table for Weighted AUC Measures(`table_wr`)
If your study involves weighting time spent in different health states (e.g., partial vs. full recovery, mild vs. severe symptoms), use the weighted functions:
```{r}
wr_table <- table_wr(
  data1 = res_list$trt1.death,
  data2 = res_list$trt0.death,
  eta   = 2,
  tau   = 15
)
wr_table
```

Again, you’ll get a 3×4 matrix showing group estimates and the difference with confidence intervals and p-value. If you want to analyze “discharge” as the event, set `eta` = 1 and pass `res_list$trt1.discharge` / `res_list$trt0.discharge`.

## Reference and Code Source
Jiyang Wen, Chen Hu, Mei-Cheng Wang, Joint Inference for Competing Risks Data Using Multiple Endpoints, Biometrics, Volume 79, Issue 3, September 2023, Pages 1635–1645, https://doi.org/10.1111/biom.13752

## Contributing
### Bug Reports: 
Submit an issue via GitHub Issues
### Contact: 
Wenqing Zhang <wzhan115@jh.edu>


## Quick Start
#### Install
```{r}
devtools::install_github("cathyzzzhang/jointCompRisk")
```

#### Load
```{r}
library(jointCompRisk)
```

#### Prepare Data
```{r}
data_list <- prepare_actt_data()
```

#### Standard CIF
```{r}
table1(data_list$trt1, data_list$trt0, tau=15, a=0, b=1, c=0)
```

#### Weighted CIF
```{r}
table_wr(data_list$trt1.death, data_list$trt0.death, eta=2, tau=15)
```

