---
title: "Competing Risks Survival Analysis with jointCompRisk"
subtitle: "Methods, Implementation, and Applications"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Competing Risks Survival Analysis with jointCompRisk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
Author: Wenqing Zhang
package: jointCompRisk
version: "`r packageVersion('jointCompRisk')`"
bibliography: refs.bib

---

## 1. Introduction
In clinical studies, patients often face multiple mutually exclusive outcomes of equal clinical importance. Traditional survival analysis focusing on a single endpoint can provide incomplete insights, particularly when competing events like recovery and death, disease-specific versus non-disease mortality, or treatment success versus treatment-related toxicity occur simultaneously. This `jointCompRisk` package implements advanced methods for analyzing competing risks data with joint inference capabilities, providing both traditional competing risks analysis and novel severity-weighted approaches that incorporate longitudinal ordinal severity scores.

**Key Features**:
1. Joint statistical inference across multiple cumulative incidence functions
2. Severity-weighted analysis incorporating longitudinal disease measurements
3. Results in clinically meaningful time units (days gained or lost, or other time units)
4. Methods validated following Wen, Wang & Hu (2023, Statistics in Medicine)

## 2. Clinical Applications and Rationale
**Hospital-based Treatment Studies**: atients either recover (treatment success, resource availability for new patients) or experience adverse outcomes including death (treatment failure).

**Cancer Research**: Multiple disease-related events (local recurrence, distant metastasis, secondary cancers, death) may respond differently to treatment interventions.

**Cardiovascular Studies**: Various clinical events may show differential treatment responses across mortality and morbidity outcomes.

The package enables simultaneous analysis of multiple competing outcomes while respecting the clinical importance of each endpoint.

## 3. Theoretical Background
### Cumulative Incidence Functions
The **cumulative incidence function (CIF)** describes the probability of experiencing a specific event type over time while accounting for competing events. Unlike standard survival analysis that treats competing events as censored, CIF provides realistic estimates by properly incorporating all competing risks.

Mathematically, the CIF for event type $k$ is defined as  

$$
F_{k}(t) = P\bigl(T \le t,\ \delta = k \bigr) = \int_{0}^{t} S\!\bigl(u^{-}\bigr)\, d\Lambda_{k}(u)
$$
where  
- $T$ is the time to the first event of any type,  
- $\delta$ indicates the type of event that occurred,  
- $S(t) = P(T > t)$ is the overall survival function (the probability no event has occurred by time $t$),  
- $\Lambda_k(t)$ is the cumulative hazard function for event type $k$,  
- $S(u^-) = \lim_{s \uparrow u} S(s)$ denotes the left-continuous limit of the survival function at time $u$.  

This formulation shows that the CIF for event type $k$ accumulates over time as the hazard for that event type increases, weighted by the probability of surviving free of any event up to that point.

### Restricted Mean Survival Time  

The **restricted mean survival time (RMST)** up to a specified time $\tau$ represents the average survival time accumulated by that point. Unlike the median survival, RMST uses the entire survival curve up to $\tau$ and is particularly useful when the survival curve does not reach 50%.

It is defined as  

$$
\text{RMST}(\tau) = E\bigl[\min(T, \tau)\bigr] = \int_0^\tau S(t) \, dt
$$  

where $S(t)$ is the overall survival function.


### Endpoint-Specific Restricted Mean Times  
In competing risks settings, we can define endpoint-specific restricted mean times:

- **Restricted Mean Time Gained (RMTG)** for beneficial events:  
  $$
  \text{RMTG}_k(\tau) = \int_0^\tau F_k(t) \, dt
  $$

- **Restricted Mean Time Lost (RMTL)** for adverse events:  
  $$
  \text{RMTL}_k(\tau) = \int_0^\tau F_k(t) \, dt
  $$

where $F_k(t)$ is the cumulative incidence function for event type $k$.

The **net benefit** combines multiple event types using clinical weights to reflect their relative importance:

$$
\text{Net Benefit}(\tau) = \sum_k w_k \int_0^\tau F_k(t) \, dt
$$  
where $w_k$ represents clinical importance weights (positive for beneficial events, negative for adverse events).

## 4. Installation and Data Requirements
**Load the package**
```{r}
# Install package
# devtools::install_github("cathyzzzhang/jointCompRisk")
library(jointCompRisk)
```

**Sample Data**
Throughout this vignette, we use simulated data that follows the structure of Adaptive COVID-19 Treatment Trials (ACTT). The dataset represents a clinical trial of 150 patients with competing risks survival data (recovery vs. death) and longitudinal ordinal severity scores measured over a 30-day follow-up period.
1. Treatment group: 1.5× faster recovery times, 1.8× improved survival
2. Severity scores: 1-8 scale (mild to severe illness) measured at scheduled visits
3. Realistic treatment effects demonstrating significant benefits across standard and weighted analyses

### Load the sample data
```{r}
#Sample datasets included with package
main_df <- read.csv("main_df.csv")
long_df <- read.csv("long_df.csv")

#View data structure
head(main_df)
head(long_df)

#Access help documentation
?main_df
?long_df
```

`main_df` contains baseline demographics, treatment assignment, censoring indicators, and time-to-event outcomes (TimeToRecovery, TimeToDeath).
`long_df` contains longitudinal severity measurements with one row per visit per participant, enabling joint modeling of how baseline characteristics and evolving severity scores predict competing outcomes.

## Data Structure Requirements

### For standard CIF analysis: your dataset needs:
1.	Patient ID: Unique identifier for each patient
2.	Time to Recovery: Time from enrollment to recovery/discharge
3.	Time to Death: Time from enrollment to death
4.	Recovery Censoring: Indicator for recovery censoring (0=event, 1=censored)
5.	Death Censoring: Indicator for death censoring (0=event, 1=censored)
6.	Treatment: Treatment group indicator (0=control, 1=treatment)

### For Severity-Weighted CIF Analysis:
In addition to the above columns, we need an additional column
1.	Baseline Score: baseline disease severity
As well as we need an additional longitudinal dataset with
1.	Patient ID: matching the main dataset
2.	Time Point: Days since treatment starts
3.	Severity Score: Disease severity score at each time point

### Part 1: Standard CIF Analysis:
Standard CIF analysis provides traditional competing risks analysis with joint inference capabilities.

#### Step 1: Data preparation 

The prep_data_cif() function preprocesses your data by excluding patients with zero survival time, handling "discharge-to-die" cases (recoded as censored for recovery), creating event time variables (etime, estatus, etype2), and splitting data by treatment group.

Customize column names to match your dataset:

```{r}
mydata_std <- prep_data_cif(
  data             = main_df,
  ID               = "ID",                               #change this
  TimeToRecovery   = "TimeToRecovery",                   #change this
  TimeToDeath      = "TimeToDeath",                      #change this
  Recov_Censoring  = "RecoveryCensoringIndicator",       #change this
  Death_Censoring  = "DeathCensoringIndicator",          #change this
  Treatment        = "Treatment"                         #change this
)
```

#### Step 2: Analysis
The `do_cif_analysis()` function performs three analyses: RMTG (time gained through recovery), RMTL (time lost through death), and Net Benefit (RMTG minus RMTL). The tau parameter sets the time horizon (e.g., tau=15 analyzes the first 15 days).

Each analysis returns a 3×4 matrix containing estimates, confidence intervals, and p-values for both treatment groups and their difference, providing comprehensive statistical inference for all competing risk endpoints simultaneously.
```{r}
res_std <- do_cif_analysis(mydata_std, tau=15)

res_std$RMGT  # Restricted Mean Time Gained
res_std$RMLT  # Restricted Mean Time Lost  
res_std$Net   # Net Benefit
```
According to the results above, using the simulated example data, 
From the RMTG table: Treatment group gained 2.59 more days of recovery time over 15 days (3.76 vs 1.17 days, p < 0.001)
RMTL: Treatment group lost 1.66 fewer days to death over 15 days (1.08 vs 2.74 days, p = 0.004)
Net Benefit: Treatment provides 4.25 net days of benefit over 15 days (2.68 vs -1.58 days, p < 0.001)

### Part 2: Weighted CIF Analysis:
Weighted analysis incorporates disease severity changes over time, providing insights when patient conditions vary during follow-up.


#### Step 1: Data Preparation

The function merges time-to-event data with longitudinal severity measurements, calculating time spent in different severity states and applying clinical weights. Severity scores are flexible (example uses scores 4-7, but any range can be used to meet your dataset's needs) with matching weights required for each state. In practice, death weights often increase with severity (State 4: 2.0, State 5: 1.5, State 6: 1.0, State 7: 0.5) to emphasize concerning states, while discharge weights often decrease with severity (State 4: 0.5, State 5: 1.0, State 6: 1.5, State 7: 2.0) to emphasize favorable recovery states. This produces separate datasets enabling computation of weighted restricted mean lifetime (WRMLT) and weighted restricted mean gain time (WRMGT). Weights can also be customized based on stakeholder priorities—hospital administrators might prioritize shorter stays, while patients might weight quality of life differently than clinical severity.

```{r}
prepped_w <- prep_data_weighted_cif2(
  data_main = main_df,
  data_long = long_df,

  wID_main              = "ID",
  wTimeToRecovery_main  = "TimeToRecovery",
  wTimeToDeath_main     = "TimeToDeath",
  wRecov_Censoring_main = "RecoveryCensoringIndicator",
  wDeath_Censoring_main = "DeathCensoringIndicator",
  wTreatment_main       = "Treatment",
  wBaselineScore_main   = "BaselineScore",
  
  wID_long              = "PersonID",
  wADY_long             = "RelativeDay",
  wScore_long           = "OrdinalScore",

  wStates_death         = c(4,5,6,7,8), 
  wWeights_death        = c(2,1.5,1,0.5,0.2),
  wStates_discharge     = c(4,5,6,7,8),
  wWeights_discharge    = c(0.2,0.5,1,1.5,2)
)
```


#### Step 2: Analysis
The function `do_weighted_cif_analysis()`performs weighted competing risks analysis using the datasets prepared by prep_data_weighted_cif(), computing Weighted RMTL (death analysis with severity-weighted time) and Weighted RMGT (discharge analysis with recovery-weighted time) at the specified time horizon tau. Unlike standard analysis that treats all time equally, this approach emphasizes clinically relevant periods - time spent in severe states before death receives higher weight in WRMTL calculations, while time spent in better states before discharge receives higher weight in WRMGT calculations. The function returns 3×4 matrices for both weighted measures, containing estimates, confidence intervals, and p-values that reflect the severity-adjustsed treatment effects on competing outcomes.

Different `tau` values reveal treatment effects over time: short-term (e.g., tau=15) shows immediate treatment effects, long-term (e.g., tau=29) reveals sustained or delayed benefits, and multiple horizons demonstrate the evolution of treatment effects. Treatment benefits may be immediate, delayed, or diminishing, so analyzing multiple time points provides comprehensive insights.

For standard analysis, `RMTG > 0` indicates more time gained through recovery, `RMTL < 0` shows less time lost to death, and `Net Benefit > 0` demonstrates overall favorable treatment effect. For weighted analysis, WRMGT represents recovery time weighted by clinical improvement, WRMTL shows death time weighted by severity progression, and both capture timing and severity of clinical outcomes. Results provide clinically meaningful metrics in time units, facilitating evidence-based treatment decisions.
```{r}
#Analysis at 15 days
res_w15 <- do_weighted_cif_analysis(prepped_w, tau=15)
res_w15$WRMLT
res_w15$WRMGT

#Analysis at 29 days
res_w29 <- do_weighted_cif_analysis(prepped_w, tau=29)
res_w29$WRMLT
res_w29$WRMGT
```
At 15 days:
WRMTL: Treatment group lost 1.34 fewer severity-weighted days (1.73 vs 3.08 days, p = 0.124)
WRMGT: Treatment group gained 2.88 more recovery-weighted days (3.83 vs 0.95 days, p < 0.001)

At 29 days:
WRMTL: Treatment group lost 3.40 fewer severity-weighted days (5.09 vs 8.49 days, p = 0.074)
WRMGT: Treatment group gained 7.10 more recovery-weighted days (10.56 vs 3.45 days, p < 0.001)

